\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2015-07-22 tex/discussion.tex
change some usages of components to variables

%!TEX root = ../paper.tex
% Our development suggests two predictions that could be used to test the theory described here. One is that,
\section{Conclusion}
Our analysis predicts that, in general, an ensemble of systems where robustness has been the predominant object of selection and has been positively selected over a sufficiently long period of time should exhibit a bias toward more hierarchical network topologies. Given the manner in which we define robustness in \ref{eq:robustnessfromstoch}, this is a very general constraint. In the short term, this prediction may be further evaluated at the levels of both metabolic and transcription factor networks, which have already been shown to display hierarchical structure, but whose dynamics have not been sufficiently well characterized to ascertain their dynamical robustness as we have defined it here \cite{Zhao2006,Bhardwaj2010,Colm,Prill2005}. Indeed, Figure 2 from \cite{Prill2005} demonstrates an enrichment of various hierarchical network architectures consisting of multiple SCCs relative to those containing cycles with single SCCs for networks with three and four components in \emph{E. coli}, \emph{S. cerevisiae}, and \emph{D. melanogaster} transcription factor networks as well as the signal transduction knowledge environment network and the \emph{C. elegans} neural network. The network architecture that we predict to be enriched by selection for robustness for three variable networks ranks 4th out of 13 in overall abundance across these networks and within each network individually with the exception of the \emph{C. elegans} neural network, and the three more abundant network architectures are all proper subgraphs of this one. These results are essentially consistent with our prediction, though, of course, we would not claim that there could not be various alternative explanations for these observations. At the ecological level, a system subjected to the environmental stress of overfishing, which may imply selection for robustness, has been observed to exhibit such a bias toward more hierarchical network architectures \cite{Bascompte2005}.
%The other prediction of the theory outlined here that may be able to be checked is the converse statement that an ensemble of systems demonstrating a positive or negative bias in their network topology toward hierarchy may have been under selection for, or respectively against, robustness. Of course, in this case, factors other than selection for robustness would not be ruled out by this analysis alone.
In the long term, this prediction may be evaluated using experimental evolution by comparing the degree of hierarchy that emerges in the evolution of gene regulatory network topology in the context of both static and fluctuating environments that impose differential selection strengths for dynamical robustness \cite{Leroi1994}.

>>>>

%!TEX root = ../paper.tex
% Our development suggests two predictions that could be used to test the theory described here. One is that,
\section{Conclusion}
Our analysis predicts that, in general, an ensemble of systems where robustness has been the predominant object of selection and has been positively selected over a sufficiently long period of time should exhibit a bias toward more hierarchical network topologies. Given the manner in which we define robustness in \ref{eq:robustnessfromstoch}, this is a very general constraint. In the short term, this prediction may be further evaluated at the levels of both metabolic and transcription factor networks, which have already been shown to display hierarchical structure, but whose dynamics have not been sufficiently well characterized to ascertain their dynamical robustness as we have defined it here \cite{Zhao2006,Bhardwaj2010,Colm,Prill2005}. Indeed, Figure 2 from \cite{Prill2005} demonstrates an enrichment of various hierarchical network architectures consisting of multiple SCCs relative to those containing cycles with single SCCs for networks with three and four variables in \emph{E. coli}, \emph{S. cerevisiae}, and \emph{D. melanogaster} transcription factor networks as well as the signal transduction knowledge environment network and the \emph{C. elegans} neural network. The network architecture that we predict to be enriched by selection for robustness for three variable networks ranks 4th out of 13 in overall abundance across these networks and within each network individually with the exception of the \emph{C. elegans} neural network, and the three more abundant network architectures are all proper subgraphs of this one. These results are essentially consistent with our prediction, though, of course, we would not claim that there could not be various alternative explanations for these observations. At the ecological level, a system subjected to the environmental stress of overfishing, which may imply selection for robustness, has been observed to exhibit such a bias toward more hierarchical network architectures \cite{Bascompte2005}.
%The other prediction of the theory outlined here that may be able to be checked is the converse statement that an ensemble of systems demonstrating a positive or negative bias in their network topology toward hierarchy may have been under selection for, or respectively against, robustness. Of course, in this case, factors other than selection for robustness would not be ruled out by this analysis alone.
In the long term, this prediction may be evaluated using experimental evolution by comparing the degree of hierarchy that emerges in the evolution of gene regulatory network topology in the context of both static and fluctuating environments that impose differential selection strengths for dynamical robustness \cite{Leroi1994}.

### 2015-07-22 tex/figuressupp.tex
change some usages of components to variables

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.9\columnwidth]{fig/jacobianvectorfields.pdf}
\caption{{\bf Vector fields resulting from the random sampling of two component systems.} System parameters of \ref{eq:arbitraryjacobianreactionnetwork} are rescaled to ensure the fixed point is located at $(1,1)$. The color of the dot located at the fixed point indicates whether it is stable (black) or unstable (gray).}
\label{fig:jacobianvectorfields}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.6\columnwidth]{fig/apstab3x3.pdf}
\caption{{\bf System stability as a function of connectivity.} The probability of stability is plotted versus connectivity for all three component systems. We note that, in contrast to robustness, the stability ranges over less than 10\% of the possible dynamic range and demonstrates no obvious significant correlation with connectivity. The red points represent the average of system stability at each connectivity.}
\label{fig:apstab3x3}
\end{figure}

>>>>

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.9\columnwidth]{fig/jacobianvectorfields.pdf}
\caption{{\bf Vector fields resulting from the random sampling of two variable systems.} System parameters of \ref{eq:arbitraryjacobianreactionnetwork} are rescaled to ensure the fixed point is located at $(1,1)$. The color of the dot located at the fixed point indicates whether it is stable (black) or unstable (gray).}
\label{fig:jacobianvectorfields}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.6\columnwidth]{fig/apstab3x3.pdf}
\caption{{\bf System stability as a function of connectivity.} The probability of stability is plotted versus connectivity for all three variable systems. We note that, in contrast to robustness, the stability ranges over less than 10\% of the possible dynamic range and demonstrates no obvious significant correlation with connectivity. The red points represent the average of system stability at each connectivity.}
\label{fig:apstab3x3}
\end{figure}

### 2015-07-22 tex/introduction.tex
change some usages of components to variables

Recent work has demonstrated that as a result of the existence of largely insensitive directions within the parameter space for such models, the approach outlined above often allows for a large variety of models to fit equivalent data \cite{Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In addition, there is often uncertainty about the very structure of such networks. Since the evolutionary process results in modifications to the underlying model this fact may be used to characterize evolutionarily effective versus neutral spaces. In this context, it is crucial to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems. This is necessary to understand in order to determine whether or not a given dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}. This can be achieved using a method common in statistical physics involving the consideration of an ensemble of systems \cite{Bialek2012}. For the case considered here, the ensemble of interest is defined by dynamical systems that, in comparison to one another, appear to have components that are randomly interlinked.
% \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}

@@

Here we determine the relationship between network hierarchy, a topological property, and \emph{dynamical robustness}, a property of the dynamics. Robustness is of interest in biological systems at all scales, and has been previously studied in the context of biochemical networks \cite{Alon1999,Shinar2010}, gene-regulatory networks \cite{VanNimwegen1999,Siegal2002,Draghi2010,Wagner2013}, and ecological networks \cite{Rohr2014}. Over physiological timescales, the robustness of a particular network state may be evaluated by determining its linear stability \cite{Davis1962}. Network states that are linearly stable, are robust to perturbations in the states of their components. For example, in the case of gene-regulatory networks, a state vector containing protein molecule counts or concentrations of each gene that is stable with respect to the dynamics of the network will exponentially suppress any relatively small modifications to the state of a gene and return to the initial stable state.

>>>>

Recent work has demonstrated that as a result of the existence of largely insensitive directions within the parameter space for such models, the approach outlined above often allows for a large variety of models to fit equivalent data \cite{Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In addition, there is often uncertainty about the very structure of such networks. Since the evolutionary process results in modifications to the underlying model this fact may be used to characterize evolutionarily effective versus neutral spaces. In this context, it is crucial to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems. This is necessary to understand in order to determine whether or not a given dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}. This can be achieved using a method common in statistical physics involving the consideration of an ensemble of systems \cite{Bialek2012}. For the case considered here, the ensemble of interest is defined by dynamical systems that, in comparison to one another, appear to have variables that are randomly interlinked.
% \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}

@@

Here we determine the relationship between network hierarchy, a topological property, and \emph{dynamical robustness}, a property of the dynamics. Robustness is of interest in biological systems at all scales, and has been previously studied in the context of biochemical networks \cite{Alon1999,Shinar2010}, gene-regulatory networks \cite{VanNimwegen1999,Siegal2002,Draghi2010,Wagner2013}, and ecological networks \cite{Rohr2014}. Over physiological timescales, the robustness of a particular network state may be evaluated by determining its linear stability \cite{Davis1962}. Network states that are linearly stable, are robust to perturbations in the states of their variables. For example, in the case of gene-regulatory networks, a state vector containing protein molecule counts or concentrations of each gene that is stable with respect to the dynamics of the network will exponentially suppress any relatively small modifications to the state of a gene and return to the initial stable state.

### 2015-07-22 tex/methods.tex
change some usages of components to variables

% The quality of interest in this paper is robustness, which is related to the concept of structural stability (see \ref{sec:dynrobustandstructstab}), whose
The evaluation of \ref{eq:robustnessonjacobians} requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one or more of its defining parameters, its rate functions, or environmental constraints that restrict it to a subset of its basins of attraction. We mean to refer to perturbations to the structure of the system itself as determined by the strengths of the couplings between the components and not only to perturbations of the state vector at a given point in time. It is justified to consider resampling elements of $A$ to generate $A'$ as a proxy for resampling elements of $\vec{p}$ to produce $\vec{p}\,'$ if any matrix $A$ can be obtained for some $F_i$, $\vec{p}$ and $\vec{x}^0$. This holds for the $F_i$ defining the Lotka-Volterra model. This is due to the fact that for a specification of non-zero real numbers for the components of $\vec{x}^0$ and any real numbers for the components of $a_{ij}$, there is a choice of parameters $\vec{p}$ given by $b_{ij} = \frac{a_{ij}}{x_i^0}$ and $r_i = - \sum_{j=1}^n \frac{x_j^0}{x_i^0} a_{ij}$ that generates those particular $a_{ij}$ as the Jacobian matrix of the dynamical system. Checking this property of the domain of realizability of the Jacobian can be done for ensembles of systems other than the Lotka-Volterra ensemble. For arbitrary biochemical reaction and gene regulatory networks, this property is likely to hold so long as not too many types of transformations are constrained from possibility. For example, a simplified version of the general form of the gene regulatory network model presented in \ref{fig:biomodelexamples} center panel is given by the system
\begin{equation}
\frac{dg_i}{dt} = \sum_{j=1}^N k_{ij} g_j,
\end{equation}
with one parameter $k_{ij} \in \mathbb{R}$ for every pair $(g_i,g_j)$ of genes. The Jacobian of this system is $A_{ij} = k_{ij}$, and, therefore, sampling parameters of the model is precisely equivalent to sampling elements of the Jacobian.
% For reaction networks, we demonstrate an ensemble from which arbitrary Jacobians can arise in \ref{sec:reactionnetjacobian}.

@@

 We refer to the most hierarchical network architecture as the directed graph associated to a total ordering on the set of system components corresponding to the set of nodes, $V$, of the graph \cite{Cormen2009}. In general, a totally ordered set is a pair $(S,R)$ consisting of a set $S$ together with a total order relation $R$ on it. An example of a total ordering is the less than or equal to relation, $R \equiv \leq$, on the subset of natural numbers $S \equiv \{1,2,3\}$ given by $R \equiv \{1 \leq 1, 2 \leq 2, 3 \leq 3, 1 \leq 2, 1 \leq 3, 2 \leq 3\}$. The graph associated to this relation is equivalent to the graph shown in \reffigscc{} top and described algebraically in the preceding paragraph. More precisely, the conditions on $R$ for arbitrary elements $x,\,y,\,z \in S$ necessary for $(S,R)$ to be a totally ordered set are
\begin{enumerate}
\item If $x R y$ and $y R x$ then $x=y$ (antisymmetry)
\item If $x R y$ and $y R z$ then $x R z$ (transitivity)
\item $x R y$ or $y R x$ (totality)
\end{enumerate}
The totality condition implies $x R x$ (reflexivity) corresponding to the fact that the directed graph associated to the total ordering has, for each node, an edge whose source and target are the same node.

>>>>

% The quality of interest in this paper is robustness, which is related to the concept of structural stability (see \ref{sec:dynrobustandstructstab}), whose
The evaluation of \ref{eq:robustnessonjacobians} requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one or more of its defining parameters, its rate functions, or environmental constraints that restrict it to a subset of its basins of attraction. We mean to refer to perturbations to the structure of the system itself as determined by the strengths of the couplings between the variables and not only to perturbations of the state vector at a given point in time. It is justified to consider resampling elements of $A$ to generate $A'$ as a proxy for resampling elements of $\vec{p}$ to produce $\vec{p}\,'$ if any matrix $A$ can be obtained for some $F_i$, $\vec{p}$ and $\vec{x}^0$. This holds for the $F_i$ defining the Lotka-Volterra model. This is due to the fact that for a specification of non-zero real numbers for the components of $\vec{x}^0$ and any real numbers for the components of $a_{ij}$, there is a choice of parameters $\vec{p}$ given by $b_{ij} = \frac{a_{ij}}{x_i^0}$ and $r_i = - \sum_{j=1}^n \frac{x_j^0}{x_i^0} a_{ij}$ that generates those particular $a_{ij}$ as the Jacobian matrix of the dynamical system. Checking this property of the domain of realizability of the Jacobian can be done for ensembles of systems other than the Lotka-Volterra ensemble. For arbitrary biochemical reaction and gene regulatory networks, this property is likely to hold so long as not too many types of transformations are constrained from possibility. For example, a simplified version of the general form of the gene regulatory network model presented in \ref{fig:biomodelexamples} center panel is given by the system
\begin{equation}
\frac{dg_i}{dt} = \sum_{j=1}^N k_{ij} g_j,
\end{equation}
with one parameter $k_{ij} \in \mathbb{R}$ for every pair $(g_i,g_j)$ of genes. The Jacobian of this system is $A_{ij} = k_{ij}$, and, therefore, sampling parameters of the model is precisely equivalent to sampling elements of the Jacobian.
% For reaction networks, we demonstrate an ensemble from which arbitrary Jacobians can arise in \ref{sec:reactionnetjacobian}.

@@

 We refer to the most hierarchical network architecture as the directed graph associated to a total ordering on the set of system variables corresponding to the set of nodes, $V$, of the graph \cite{Cormen2009}. In general, a totally ordered set is a pair $(S,R)$ consisting of a set $S$ together with a total order relation $R$ on it. An example of a total ordering is the less than or equal to relation, $R \equiv \leq$, on the subset of natural numbers $S \equiv \{1,2,3\}$ given by $R \equiv \{1 \leq 1, 2 \leq 2, 3 \leq 3, 1 \leq 2, 1 \leq 3, 2 \leq 3\}$. The graph associated to this relation is equivalent to the graph shown in \reffigscc{} top and described algebraically in the preceding paragraph. More precisely, the conditions on $R$ for arbitrary elements $x,\,y,\,z \in S$ necessary for $(S,R)$ to be a totally ordered set are
\begin{enumerate}
\item If $x R y$ and $y R x$ then $x=y$ (antisymmetry)
\item If $x R y$ and $y R z$ then $x R z$ (transitivity)
\item $x R y$ or $y R x$ (totality)
\end{enumerate}
The totality condition implies $x R x$ (reflexivity) corresponding to the fact that the directed graph associated to the total ordering has, for each node, an edge whose source and target are the same node.
